---
title: "R for beginners"
author: "Alexander Matrunich"
date: "08/23/2014"
output: html_document
---

We continue on from our Hello world example, now we will not salute the world in
general, but create greetings from the most popular names in US. To complete the task we will learn how to install and load packages in R, how to deal with data frames and how to get basic statistics about a data set.

## R: core and packages 

R software is not a monolithic structure, it’s very flexible. Except for
beginners, there are no two R users who have identical installations of R. R
consists of a core which holds basic functions, and additional stand-alone
packages. When you download a R distribution from CRAN you get the core and a
bunch of base packages. Usually this basic functionality will not be enough for
your specific tasks and you will want to add some additional packages. As of November 2013 there were more than 5000 packages on CRAN, 700 on BioConductor and unknown number of packages at GitHub.

You can install the new package called "babynames" with the next command. 

```{r install_lib, eval=FALSE}
install.packages("babynames")
```

Depending on which operating system you are using, you will get specific output
about results of the installation. The package will be installed in your user
directory and will be available only for you. If you want the package to be
available for all users of the computer you have to start R-session with
root/administrative privileges and install it with the install.packages()
function. In this case the package will be placed in system directory and will
be available to all users.

So here we have installed the package named "babynames". A package consists of
functions and/or data. The particular package "babynames" consists of data sets
only: USA baby names that have been used at least five times per year, from
years 1880 to 2013. Let's look inside.

We've installed a package however to use a package during R-session we have to
tell R about this and "load" the package.

```{r attach_lib}
library("babynames")
```

After execution of this command R will know that it should look for names of
functions or data sets, inside R core and base packages and also the library "babynames".

## Discovering the data set

Let's look at the size of the babynames data set.

```{r dim_data}
# Number of rows and columns of the data set
dim(babynames)
```
Wow! It is a huge data set! It has `r dim(babynames)[1]` rows! I'm afraid it's
not possible to put this table into an Excel spreadsheet. Sorry about it!

Let's look at the top of the dataset with the headings so we get an idea of what
data is actually in the data set.

```{r head_data}
# First six rows of the data set
head(babynames)
```

Let's find the most popular name. I'm asking R to output the row where "n"
(number) is the maximum.

```{r max_n}
babynames[babynames$n == max(babynames$n),] 
```
OK, it was "Linda" in the year 1947. Let's dig into the command I used. Here we
deal with several R-instruments. Dollar sign ($), square brackets([]), logical
operators and a function (max()).

The dollar sign is used when you want to access an object inside of another
object. The data.frame babynames consists of 5 columns (vectors) of numbers. I
use babynames$n to return column n from babynames data set as a vector.

Then I put this vector of numbers into function max(), which will return the
element of that vector that is largest. The construction babynames\$n ==
max(babynames$n) returns information about which element(s) of the vector
babynames\$n are maximum. The equal sign is doubled to avoid confusion with the
assign operator.  x = 5 means 'assign the value of 5 to the variable x' while x
== 5 means 'check and see if the value contained in variable x is exactly 5'.

Square brackets are a brother of the dollar sign: they also allow user to
extract specific parts of object. babynames[A, B] means 'please return from data
set babynames, rows which equal A, and columns which equal B'. So here I asked R
to return from babynames data set, rows where n is the maximum, and all the columns
where the vale of n is the maximum.

Ooops, error in my logic! In 1947 there were `r babynames$n[babynames$n == max(babynames$n)]` children named Linda.  The number is the maximum in the data set. However it doesn't mean that Linda was the most
popular name because the total number of children per year differs. We should
look at prop (proportion) column which indicates proportional share of children
given that name in that year.

```{r max_prop}
babynames[babynames$prop == max(babynames$prop),]
```

OK, it's John from 1880. 8% of children born in 1880 got name John.

Let's expand our task and print Hello to the most popular names of girls born
from 1903 untill 1911. I'll show you more advanced approach with using of
another package used for data manipulation. It's called dplyr, that stands for "Data.frame's apPLY in R", where apply means to apply different functions.


```{r install_dplyr, eval=FALSE}
install.packages("dplyr") # You already know what does it mean
```

```{r attach_dplyr, message=FALSE}
library("dplyr")
```

I divide our task in two steps. First we get the names and put them into a variable, secondly we print the desired strings.

```{r selecting_names}
girls_names <- babynames %>%   # We work with babynames data.frame
                               # and put results into girls_names
  filter(year %in% 1903:1911 & # Filtering babynames by years
           sex == 'F') %>%     # and by gender
  group_by(year) %>%           # We want results per each year
  filter(prop == max(prop))    # We keep rows with max value of prop
```

OK, we've got the most popular girls' names from 1903 to 1911, now let's salute them.

```{r print_names}
paste("Hello ", girls_names$name, "!", sep = "")
```

Mary in all cases?! Let's check our data.frame:

```{r print_df}
girls_names
```

Yes, it seems like Mary was the most popular girl name in USA in the beginning of XXth century.
